% Xavier Danaux (xdanaux@gmail.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
% Important note:
% This template requires the moderncv.cls and .sty files to be in the same
% directory as this .tex file. These files provide the resume style and themes
% used for structuring the document.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%    PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[11pt,a4paper,sans]{moderncv} % Font sizes: 10, 11, or 12; paper sizes: a4paper, letterpaper, a5paper, legalpaper, executivepaper or landscape; font families: sans or roman

\renewcommand{\refname}{References}

\moderncvstyle{casual} % CV theme - options include: 'casual' (default), 'classic', 'oldstyle' and 'banking'
\moderncvcolor{grey} % CV color - options include: 'blue' (default), 'orange', 'green', 'red', 'purple', 'grey' and 'black'

\usepackage[numbers,sort&compress]{natbib}
\usepackage{lipsum} % Used for inserting dummy 'Lorem ipsum' text into the template
\usepackage[scale=0.78]{geometry} % Reduce document margins
\usepackage{xspace}

\bibliographystyle{unsrtnat}

%----------------------------------------------------------------------------------------
%    NAME AND CONTACT INFORMATION SECTION
%----------------------------------------------------------------------------------------
\newcommand{\ra}{read-alignment\xspace}
\newcommand{\qm}{quasi-mapping\xspace}
\newcommand{\rapmap}{\texttt{RapMap}\xspace}
\newcommand{\alevin}{\texttt{alevin}\xspace}
\newcommand{\Alevin}{\texttt{Alevin}\xspace}

\firstname{Research Statement} % Your first name
\familyname{} % Your last name
\title{Avi Srivastava | \today}
\address{277 Sheep Pasture}{East Setauket-11733, New York}
\mobile{(631)974-8087}

\email{asrivastava@cs.stonybrook.edu}


%----------------------------------------------------------------------------------------

\begin{document}
\makecvtitle % Print the CV title

My main research interest is to a better understand the dynamic gene expression landscapes through efficient algorithms and statistical
inference approaches that account for uncertainty. This broad goal has led me into the field of Computational Biology and the study of RNA Sequencing (RNA-seq) experiments, in particular.

\section{Published Research}
RNA-seq technologies are evolving rapidly, and with them, the requirement for fast and accurate tools to determine and analyze the origin of reads in the generated data. 
The first step of many RNA-seq analyses requires to solve the problem of \ra. When reads are aligned to a collection of reference sequences that share a 
substantial amount of sub-sequence (near or exact repeats), a single read can have many potential alignments, and considering all such alignments can be crucial for downstream analysis. 

{\hskip 2em}I introduced a novel concept, \qm, and an efficient algorithm implementing this approach called \rapmap which maps RNA-seq reads (sequences) to the reference sequence(s) (transcriptome). 
\rapmap is capable of \textit{mapping} sequencing reads to a target transcriptome substantially faster than existing alignment tools. The algorithm I employed to implement \qm uses several efficient 
data structures and takes advantage of the special structure of shared sequence prevalent in transcriptomes to rapidly provide highly-accurate mapping information. \rapmap is
well accepted by the community, and to date, the manuscript has been downloaded $\sim$9k times and cited 44 times \cite{rxivist}.

{\hskip 2em}While bulk RNA-seq is an established method to perform genome-wide quantification\cite{rseq},
bulk experiments average-out the expression patterns of individual cells 
(or cell types) across millions of cells, losing cell-level heterogeneity which is crucial to understand
the gene expression landscape. Moreover, quantification tools for bulk RNA-seq cannot be directly used for droplet based single-cell RNA-seq (dscRNA-seq) data\cite{dropseq, indrop, tenx}. 
I introduced \alevin, a fast end-to-end pipeline to process dscRNA-seq data, addressing
the inherent bias in existing tools which discard gene-ambiguous reads and improving the
accuracy of gene abundance estimates. \Alevin is considerably faster, typically 8 times faster,
than existing gene-quantification approaches, while also using less memory.

\section{Future Research}
Given the success of both \rapmap and \alevin, I continued pursuing this line of research and collaborated to develop other downstream tools \cite{rapclust, ddfact, pufferfish, swish, minnow}. 
In addition, I plan to generalize both methods to make them more robust and extend their capabilities. 
While \qm is a useful novel concept, it trades-off some accuracy for speed. This does not have a huge
impact on typical RNA-seq experiments. However, in \textit{relatively} complex datasets, the loss in accuracy can be significant. In ongoing work, I have updated \rapmap to employ alignments to improve 
quantifications in an efficient, dynamic (dataset dependent) way. This optimizes the
speed/accuracy tradeoff, making the estimates almost as accurate as end-to-end alignment-based
methods, while still using resources similar to \rapmap. Moreover, as \alevin internally uses \rapmap for transcriptome mapping, I believe the improvement in mapping quality will be directly reflected in 
the accuracy of \alevin, while also improving the performance of other downstream tools.

\begin{thebibliography}{1}
\bibitem{rapclust} Srivastava, Avi, et al. 
"Accurate, fast and lightweight clustering of de novo transcriptomes using fragment equivalence classes." 
arXiv preprint arXiv:1604.03250 (2016).

\bibitem{dropseq} Macosko, E. Z., Basu, A., Satija, R., Nemesh, J., Shekhar, K., Goldman, M., ... \& Trombetta, J. J. (2015). 
"Highly parallel genome-wide expression profiling of individual cells using nanoliter droplets". 
Cell, 161(5), 1202-1214.

\bibitem{ddfact} Zakeri, M., Srivastava, A., Almodaresi, F., \& Patro, R. (2017). 
"Improved data-driven likelihood factorizations for transcript abundance estimation". 
Bioinformatics, 33(14), i142-i151.

\bibitem{indrop} Klein, A. M., Mazutis, L., Akartuna, I., Tallapragada, N., Veres, A., Li, V., ... \& Kirschner, M. W. (2015). 
"Droplet barcoding for single-cell transcriptomics applied to embryonic stem cells". 
Cell, 161(5), 1187-1201.

\bibitem{pufferfish} Almodaresi, F., Sarkar, H., Srivastava, A., \& Patro, R. (2018). 
"A space and time-efficient index for the compacted colored de Bruijn graph".
Bioinformatics, 34(13), i169-i177.

\bibitem{swish} Zhu, A., Srivastava, A., Ibrahim, J. G., Patro, R., \& Love, M. I. (2019). 
"Nonparametric expression analysis using inferential replicate counts". BioRxiv.

\bibitem{tenx} Zheng, G. X., Terry, J. M., Belgrader, P., Ryvkin, P., Bent, Z. W., Wilson, R., ... \& Gregory, M. T. (2017). 
"Massively parallel digital transcriptional profiling of single cells". Nature communications, 8, 14049.

\bibitem{minnow} Sarkar, H., Srivastava A., Patro R. (2019)
"Minnow: A principled framework for rapid simulation of dscRNA-seq data at the read level"
ISMB-2019

\bibitem{rseq} Mortazavi, A., Williams, B. A., McCue, K., Schaeffer, L., \& Wold, B. (2008). 
"Mapping and quantifying mammalian transcriptomes by RNA-Seq". 
Nature methods, 5(7), 621.

\bibitem{rxivist} {rxivist: webpage} 
{\href{https://rxivist.org/papers/5324} https://rxivist.org/papers/5324}

\end{thebibliography}

\end{document}
